---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CensRCalc

<!-- badges: start -->
<!-- badges: end -->

```{r, echo = FALSE, results = "asis", comment=""}
descr <- packageDescription("CensRCalc")
cat(descr$Title)
cat("\n\n")
cat(descr$Description)
```

The documentation is available at 
[https://iden-project-uas-darmstadt.github.io/CensRCalc/](
  https://iden-project-uas-darmstadt.github.io/CensRCalc/)

## Installation

You can install the development version of CensRCalc like so:

``` r
# With remotes
remotes::install_github("IDEN-Project-UAS-Darmstadt/CensRCalc")
# With pak (recommended for speed)
pak::pkg_install("IDEN-Project-UAS-Darmstadt/CensRCalc")
```

Releases of the library can be found 
[here](https://github.com/IDEN-Project-UAS-Darmstadt/CensRCalc/releases).

## Examples

This package provides functions to estimate the parameter of the random
censoring distribution based on a given event time distribution and
covariate distributions.

```{r}
library(CensRCalc)

event_rate <- function(x, trt) exp(-0.5 + 0.3 * x + 0.2 * trt)
event_surv <- function(t, x, trt) 1 - pexp(t, rate = event_rate(x, trt))
cov_dens <- function(x, trt) dnorm(x, mean = 0, sd = 1) * 1 / 2
cov_bounds <- list(x = c(-Inf, Inf), trt = list(0, 1))

# Default is a exponential censoring distribution
estimator <- estimate_cens_prop(event_surv, cov_dens, cov_bounds)

# We can now estimate the censoring prop. with different lambda_c values
sapply(c(0.5, 0.7), estimator)
```

```{r}
print(estimator)
plot(estimator)
```

We can also find a `lambda_c` that results in a desired censoring proportion.

```{r}
progressr::with_progress({ # Progress bar is optional
  lambda_c_50 <- find_cens_param(0.5, event_surv, cov_dens, cov_bounds)
  print(lambda_c_50)
})
```


# Development

This package comes with a Devcontainer that can be used to develop the package
in a reproducible environment.

Use anything available in the `devtools` package to develop the package.

```r
library(devtools)
document() # to update documentation and roxygen functionality
load_all() # to load the package functions for development
build_readme() # to update the README
build_vignettes() # to build the vignettes (deprecated, but still works)
build_site() # to build the pkgdown site
build_manual(path=".") # to build the PDF manual
test() # to run tests
check() # to check the package
covr::package_coverage() # to check code coverage
print(covr::report(browse=F))
styler::style_pkg() # to style the code
lint() # to check the code for linting issues
```

## Releasing a new version

Releases and pre-releases are managed automatically via GitHub Actions whenever
changes are pushed to `main` or `dev`.

To publish a new official release:

1. **Prepare on `dev`:** Update the version number in `DESCRIPTION`
  (e.g., `0.2.0`) and document all changes in `NEWS.md` under a corresponding
  version heading.
2. **Rebuild README:** Run `devtools::build_readme()` to reflect any changes.
3. **Merge to `main`:** Open a Pull Request from `dev` to `main` and merge
  it after checks pass.
4. **Automated Pipeline:** The `release.yaml` workflow triggers on push to
  `main` and will:
   * Execute package checks and build the package.
   * Generate the package tarball (`.tar.gz`) and LaTeX PDF Manual (`art/Manual.pdf`).
   * Tag the commit (e.g., `v0.2.0`) and extract release notes from `NEWS.md`.
   * Publish a new GitHub Release with attached build assets.
5. **Reset `dev`:** Switch back to `dev` and bump to a development version:

```bash
git checkout dev
git pull origin main
Rscript -e "usethis::use_dev_version()"
git commit -am "Bump to development version"
git push origin dev
```
