---
description: "Examples for common usage of CensRCalc functions."
title: "Examples"
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    rmarkdown::pdf_document:
        toc: true
        number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="CensRCalc")`'
link-citations: true
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
cur_def_bounds <- formals(CensRCalc::find_cens_param)$target_bounds

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  dpi = 100,
  out.width = "100%",
  warning = FALSE
)
```

This vignette provides examples for common usage of functions in the
`help(CensRCalc)` package.

For each situation we need to define the event time distribution and
the random censoring distribution. When calling the functions, we need to 
specify the survival function  $S_{T^*\mid\mathbf{X}}$ of the event time 
(conditional on the covariates, if there are any), the survival and density 
functions $S_{C\mid\lambda_{C_{\text{rnd}}}}$ and $f_{C\mid\lambda_{C_{\text{rnd}}}}$ of the 
censoring time and, if applicable, the joint density function $f_\mathbf{X}$ and 
bounds of the covariates.

For each example, we want to achieve a expected 25% proportion of censoring.

```{r common}
library(CensRCalc)

set.seed(1234)
pc_target <- 0.25
```

## Weibull with Exponential Censoring

The event time follows a Weibull distribution with shape parameter $\alpha = 2$
and scale parameter $\lambda_{T^*} = 10$, that is 
$$
T^*\sim \text{Weibull}(\alpha, \lambda_{T^*}).
$$

The random censoring time follows an exponential distribution with rate parameter
$\lambda_{C_{rnd}}$, that is
$$
C_{\text{rnd}}\sim \text{Exp}(\lambda_{C_{rnd}})
$$ 
leading to $S_{C_{\text{rnd}}}(t) = \exp(-\lambda_{C_{rnd}} t)$ and 
$f_{C_{\text{rnd}}}(t) = \lambda_{C_{rnd}} \exp(-\lambda_{C_{rnd}} t)$.

The default of the package is to use exponential censoring, so we only need
to specify the survival function of the event time.

The default bounds for the $\lambda_{C_{rnd}}$ parameter are `r cur_def_bounds[1]`
and `r cur_def_bounds[2]`.

```{r weibull_exp_no_cov}
lambda_evt <- 0.1
shape <- 2

event_surv <- \(t) 1 - pweibull(t, shape = shape, scale = 1 / lambda_evt)

lambda_c_50 <- find_cens_param(pc_target, event_surv)
lambda_c_50
```

We can now simulate data. 

```{r weibull_exp_no_cov_val}
n <- 10**6

event_times <- rweibull(n, shape = shape, scale = 1 / lambda_evt)
censoring_times <- rexp(n, rate = lambda_c_50$parameter)

observed_times <- pmin(event_times, censoring_times)
censoring_indicators <- as.integer(event_times <= censoring_times)
mean(censoring_indicators == 0)
```

## Weibull with Exponential Censoring and Covariates

The event time follows a Weibull distribution with shape parameter $\alpha = 2$
and scale parameter 
$\lambda_{T^*} = 10*\exp(-0.5 * x1 + 0.3 * x2 * \text{trt} - 0.2 * \text{trt})$,
that is
$$
T^*\mid\mathbf{X}\sim \text{Weibull}(\alpha, \lambda_{T^*}).
$$

The random censoring time follows an exponential distribution with rate parameter
$\lambda_{C_{rnd}}$, that is
$$
C_{\text{rnd}}\sim\text{Exp}(\lambda_{C_{rnd}}).
$$

We have three covariates: A standard normal covariate $X_1\sim\text{N}(0,1)$, 
a uniform covariate $X_2 \sim \text{U}(0,1)$ and a binary treatment indicator 
$\text{trt}\sim\text{B}(0.5)$.

```{r weibull_exp_cov}
lambda_evt <- \(x1, x2, trt) 0.1 * exp((-0.5 * x1 + 0.3 * x2 * trt - 0.2 * trt))
cov_dens <- \(x1, x2, trt) dnorm(x1) * dunif(x2, 0, 1) * dbinom(trt, 1, 0.5)

# Note that we give the bounds for trt as a list,
# because it is a discrete variable.

cov_bounds <- list(
  x1 = c(-Inf, Inf),
  x2 = c(0, 1),
  trt = list(0, 1)
)
shape <- 2

event_surv <- function(t, x1, x2, trt) {
  1 - pweibull(t, shape = shape, scale = 1 / lambda_evt(x1, x2, trt))
}

# This will take a bit longer to compute
lambda_c_50 <- find_cens_param(pc_target, event_surv,
  covariate_density = cov_dens,
  covariate_bounds = cov_bounds
)
lambda_c_50
```

The warnings are caused by the numerical integration where the bounds $(-\infty, \infty)$ 
of covariate $X_1$ are inserted. This does not affect the result. 

We can now simulate data. 

```{r weibull_exp_cov_val}
n <- 10**6

cov_data <- data.frame(
  x1 = rnorm(n),
  x2 = runif(n, 0, 1),
  trt = rbinom(n, 1, 0.5)
)
event_times <- rweibull(
  n,
  shape = shape,
  scale = 1 / lambda_evt(
    cov_data$x1,
    cov_data$x2,
    cov_data$trt
  )
)
censoring_times <- rexp(n, rate = lambda_c_50$parameter)

observed_times <- pmin(event_times, censoring_times)
censoring_indicators <- as.integer(event_times <= censoring_times)
mean(censoring_indicators == 0)
```

## Lognormal with Lognormal Censoring


We first use `CensRCalc::estimate_cens_prop()` to see the effect of different
censoring parameters on the proportion of censoring.

We have a lognormal distributed event time with parameters $\mu_{T^*} = 6$ and 
$\sigma_{T^*} = 1.1$, that is

$$
T^* \sim \text{LN}(\mu_{T^*}, \sigma_{T^*}) \Leftrightarrow 
\ln(T^*) \sim \text{N}(\mu_{T^*}, \sigma_{T^*}).
$$

Also the random censoring is lognormal distributed with fixed 
$\sigma_C = 1.2$ and $\mu_C = \lambda_C$ is the parameter of interest, that is
$$
C \sim \text{LN}(\lambda_C, \sigma_C).
$$

```{r logn_logn_no_cov}
event_surv <- \(t) 1 - plnorm(t, meanlog = 6, sdlog = 1.1)
cens_sdlog <- 1.2
cens_surv <- \(t, mean) 1 - plnorm(t, meanlog = mean, sdlog = cens_sdlog)
cens_dens <- \(t, mean) dlnorm(t, meanlog = mean, sdlog = cens_sdlog)

bounds <- c(0.1, 10)

est_fun <- estimate_cens_prop(event_surv,
  cens_survival = cens_surv,
  cens_density = cens_dens,
  target_bounds = bounds,
  target = "mean"
)

x <- seq(bounds[1], bounds[2], length.out = 50)
y <- sapply(x, est_fun)
mean_c_50 <- find_cens_param(pc_target, event_surv,
  cens_survival = cens_surv,
  cens_density = cens_dens,
  target = "mean",
  target_bounds = bounds
)

plot(
  x,
  y,
  type = "l",
  xlab = "Mean of log(Censoring Time)",
  ylab = "Proportion Censored"
)
points(mean_c_50$parameter, pc_target, col = "red", pch = 19)
```

We can now simulate data. 

```{r logn_logn_no_cov_val}
n <- 10**6

event_times <- rlnorm(n, meanlog = 6, sdlog = 1.1)
censoring_times <- rlnorm(n, meanlog = mean_c_50$parameter, sdlog = cens_sdlog)

observed_times <- pmin(event_times, censoring_times)
censoring_indicators <- as.integer(event_times <= censoring_times)
mean(censoring_indicators == 0)
```

## Weibull with Exponential Censoring, Covariates, Administrative Censoring and Accrual

As in a previous example, the event time follows a Weibull distribution with 
shape parameter $\alpha = 2$ and scale parameter 
$\lambda_{T^*} = 10*\exp(-0.5 * x1 + 0.3 * x2 * \text{trt} - 0.2 * \text{trt})$,
that is
$$
T^*\mid\mathbf{X}\sim \text{Weibull}(\alpha, \lambda_{T^*}).
$$

The random censoring time follows an exponential distribution with rate parameter
$\lambda_{C_{rnd}}$, that is 
$$
C_{\text{rnd}}\sim \text{Exp}(\lambda_{C_{rnd}})
$$ 
leading to $S_{C_{\text{rnd}}}(t) = \exp(-\lambda_{C_{rnd}} t)$ and 
$f_{C_{\text{rnd}}}(t) = \lambda_{C_{rnd}} \exp(-\lambda_{C_{rnd}} t)$.

We have three covariates: A standard normal covariate $X_1\sim\text {N}(0,1)$, 
a uniform covariate $X_2 \sim \text{U}(0,1)$ and a binary treatment indicator 
$\text{trt}\sim\text{B}(0.5)$.

In addition, we have an accrual time of $\tau_{\text{acc}} = 1$ and an 
administrative censoring time of $\tau_{\text{adm}} = 3$, inducing the random 
administrative censoring variable 
$$
C_{\text{adm}}\sim \text{U}(2,3)
$$ 
with
$S_{C_{\text{adm}}}(t) = 3-t$ and $f_{C_{\text{adm}}}(t) = 1$.

```{r weibull_exp_cov_acc_adm}
lambda_evt <- \(x1, x2, trt) 0.1 * exp((-0.5 * x1 + 0.3 * x2 * trt - 0.2 * trt))
cov_dens <- \(x1, x2, trt) dnorm(x1) * dunif(x2, 0, 1) * dbinom(trt, 1, 0.5)

# Note that we give the bounds for trt as a list,
# because it is a discrete variable.

cov_bounds <- list(
  x1 = c(-Inf, Inf),
  x2 = c(0, 1),
  trt = list(0, 1)
)
shape <- 2

event_surv <- function(t, x1, x2, trt) {
  1 - pweibull(t, shape = shape, scale = 1 / lambda_evt(x1, x2, trt))
}

time_acc <- 10
time_adm <- 20

lambda_c_50 <- find_cens_param(pc_target, event_surv,
  covariate_density = cov_dens,
  covariate_bounds = cov_bounds,
  time_admin_cens = time_adm,
  time_accrual = time_acc
)
lambda_c_50
```

The warnings are caused by the numerical integration where the bounds $(-\infty, \infty)$ 
of covariate $X_1$ are inserted. This does not affect the result. 

We can now simulate data. 

```{r weibull_exp_cov_acc_adm_val}
n <- 10**6

cov_data <- data.frame(
  x1 = rnorm(n),
  x2 = runif(n, 0, 1),
  trt = rbinom(n, 1, 0.5)
)
event_times <- rweibull(
  n,
  shape = shape,
  scale = 1 / lambda_evt(
    cov_data$x1,
    cov_data$x2,
    cov_data$trt
  )
)
rnd_censoring_times <- rexp(n, rate = lambda_c_50$parameter)
adm_censoring_times <- runif(n, time_adm - time_acc, time_adm)

censoring_times <- pmin(rnd_censoring_times, adm_censoring_times)
observed_times <- pmin(event_times, censoring_times)
censoring_indicators <- as.integer(event_times <= censoring_times)
mean(censoring_indicators == 0)
```
