---
description: "Mathematical and statistical details behind CensRCalc"
title: "Full Details and Usage"
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    rmarkdown::pdf_document:
        toc: true
        number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="CensRCalc")`'
link-citations: true
vignette: >
  %\VignetteIndexEntry{Full Details and Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  dpi = 100,
  out.width = "100%"
)
```

## Introduction

This vignette provides an overview of the functionalities and usage of the 
`help(CensRCalc)` package. This package is used to calculate the parameter of the
random censoring distribution so that the expected censoring proportion
in a simulated dataset is equal to a given target value. This package only covers right-censoring.

The theory behind this package is based on the work by @Wan2017.


## Mathematical Background

In this chapter we describe the mathematical background of the censoring
parameter calculation.


### Censoring without accrual {#sec:no_accrual}

First, we consider the special case where patients are accrued at the same
time, i.e. at the start of a clinical trial. We assume independent censoring,
i.e. the event time conditional on a $d$-dimensional covariate vector 
$\mathbf{X}$, that is $T^{*}\mid\mathbf{X}\sim F_{T^*\mid\mathbf{X}}$, is independent 
of the random censoring time $C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}\sim 
F_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}$ where 
$\lambda_{C_{\text{rnd}}}$ is the parameter of the random censoring 
distribution:

\begin{equation*}
    T^{*}\mid\mathbf{X} \perp\!\!\!\perp C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}
\end{equation*}

The administrative censoring time $\tau_{\text{adm}}$ is fixed. Thus, the actual
censoring time $C$ is the minimum of the random censoring time $C_{\text{rnd}}$
and the administrative censoring time $\tau_{\text{adm}}$:

\begin{equation*}
    C := \min(C_{\text{rnd}}, \tau_{\text{adm}})
\end{equation*}

Then the observed time $T$ is the minimum of the event
time $T^{*}$ and the censoring time $C$:

\begin{equation*}
    T := \min(T^{*}, C)
\end{equation*}

The event indicator is given by:
\begin{equation*}
    \delta := I(T^{*} \leq C) = \begin{cases}
        1 & \text{, } T^{*} \leq C \\
        0 & \text{, otherwise}
    \end{cases}
\end{equation*}

Let $p_{C}$ be the target value of the expected censoring proportion. The
goal is to calculate the parameter $\lambda_{C_{\text{rnd}}}$ of the random censoring
distribution $F_{C_{\text{rnd}}}$ so that the expected censoring proportion in a
simulated dataset is $p_{C}$. For this, we derive a formula for
$\lambda_{C_{\text{rnd}}}$ depending on the expected censoring proportion $p_{C}$
and the parameters of the event time distribution $F_{T^{*}}$ as well as the
administrative censoring time $\tau_{\text{adm}}$. This leads to an equation that can
be solved for $\lambda_{C_{\text{rnd}}}$ as a root finding problem.

We start with the probability of censoring for a given subject,
$P(\delta=0\mid\mathbf{X})$. Using the assumption of independent censoring 
we know that

\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= P(\{C_{\text{rnd}} \leq T^{*}, C_{\text{rnd}} \leq \tau_{\text{adm}}\} \vee \{\tau_{\text{adm}} < T^{*}, \tau_{\text{adm}} < C_{\text{rnd}}\}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) \\
                      &= P(C_{\text{rnd}} \leq T^{*}, C_{\text{rnd}} \leq \tau_{\text{adm}}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) + P(\tau_{\text{adm}} < T^{*}, \tau_{\text{adm}} < C_{\text{rnd}}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) \\
                      &= P(C_{\text{rnd}} \leq \min(T^{*}, \tau_{\text{adm}})\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) + P(\tau_{\text{adm}} < T^{*}\mid\mathbf{X})P(\tau_{\text{adm}} < C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}) \\
                      &= \!\begin{aligned}[t]
                            &\int_{0}^{\tau_{\text{adm}}} P(c \leq T^{*}\mid\mathbf{X})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c) + \\
                            &P(\tau_{\text{adm}} < T^{*}\mid\mathbf{X})P(\tau_{\text{adm}} < C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}})
                         \end{aligned} \\
                      &= \int_{0}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c)dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c) + S_{T^{*}\mid\mathbf{X}}(\tau_{\text{adm}})S_{C_{\text{rnd}}\mid \lambda_{C_{\text{rnd}}}}(\tau_{\text{adm}}) \\
                      &= \int_{0}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c)f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c)dc + S_{T^{*}\mid\mathbf{X}}(\tau_{\text{adm}})S_{C_{\text{rnd}}\mid \lambda_{C_{\text{rnd}}}}(\tau_{\text{adm}})
\end{align*}
where $S_{T^{*}\mid\mathbf{X}}$ and 
$S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}$ are the survival functions of
the event time and random censoring time conditional on the covariate vector $\mathbf{X}$ and the random censoring parameter $\lambda_{C_{\text{rnd}}}$, respectively.

The first term (the integral) describes the probability of random censoring before the administrative censoring.
The second term describes the probability that the administrative censoring happens before the event or random censoring.
Note that $P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}})$ is a random variable
which takes on values depending on the covariates $\mathbf{X}$ of a specific
subject.
Now we have to integrate $P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}})$
over the covariate distribution $F_{X}$ to get the probability of censoring in
the simulated dataset, independent of the covariates of a specific subject:

\begin{equation*}
    P(\delta = 0\mid\lambda_{C_{\text{rnd}}}) = E_{X}(P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}})) = \int_{D_{X}} P(\delta=0\mid\mathbf{X}=x, \lambda_{C_{\text{rnd}}}) dF_{X} = \int_{D_{X}} P(\delta=0\mid\mathbf{X}=x, \lambda_{C_{\text{rnd}}})f_{X}(x) dx,
\end{equation*}

where $D_{X}$ is the domain of the covariate vector $X$.
This d-dimensional integral must usually be computed numerically. The packages
`help(calculus)` and `help(cubature)` are used for this purpose.

Now we can solve for  $\lambda_{C_{\text{rnd}}}$ using a root finding algorithm. 
The function to be solved is given by:

\begin{equation*}
    P(\delta = 0\mid\lambda_{C_{\text{rnd}}}) = p_{C} \Longleftrightarrow P(\delta = 0\mid\lambda_{C_{\text{rnd}}}) - p_{C} = 0.
\end{equation*}

To solve this root finding problem we use the `stats::uniroot` function that implements a bracketing root finding algorithm.


### Censoring with accrual

In the general case, patients are accrued within a time span 
$[0, \tau_{\text{acc}}]$. We consider uniform distributed accrual times and thus, 
the random accrual time is $R \sim U([0, \tau_{\text{acc}}])$. The other parameters
and variables are identical to those in the special case. Stochastically, the
random process of accrual can be equivalently modelled by making the
administrative censoring time a random variable, $C_{\text{adm}}$, given by the
following definition:

\begin{equation*}
    C_{\text{adm}} := \tau_{\text{adm}} - R.
\end{equation*}
In other words, after accrual a patient can only be observed until the
administrative censoring time is reached. Consequently, the time a patient
can be observed after accrual, the follow-up time, is given by the difference between the
administrative censoring time and the actual accrual time of that patient.

It follows that $C_{\text{adm}} \sim U([\tau_{\text{adm}} - \tau_{\text{acc}}, \tau_{\text{adm}}])$.
Furthermore, we assume independence between $T^{*}\mid\mathbf{X}$ and $C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}$ as well
as $C_{\text{adm}}$:

\begin{equation*}
    T^{*}\mid\mathbf{X} \perp\!\!\!\perp C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}} \quad \wedge \quad T^{*}\mid\mathbf{X} \perp\!\!\!\perp C_{\text{adm}}
\end{equation*}

Again, we start with the probability of censoring for a given subject:
\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= P(\{C_{\text{rnd}} < T^{*}, C_{\text{rnd}} < C_{\text{adm}}\} \vee \{C_{\text{adm}} < T^{*}, C_{\text{adm}} < C_{\text{rnd}}\}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) \\
                                        &= P(C_{\text{rnd}} < T^{*}, C_{\text{rnd}} < C_{\text{adm}}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) + P(C_{\text{adm}} < T^{*}, C_{\text{adm}} < C_{\text{rnd}}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) \\
                                        &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}}} P(c_{\text{rnd}} < T^{*}, c_{\text{rnd}} < C_{\text{adm}}\mid\mathbf{X})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} P(c_{\text{adm}} < T^{*}, c_{\text{adm}} < C_{\text{rnd}}\mid\mathbf{X}, \lambda_{C_{\text{rnd}}})dF_{C_{\text{adm}}}(c_{\text{adm}})
                                            \end{aligned} \\
                                        &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}}} P(c_{\text{rnd}} < T^{*}\mid\mathbf{X})P(c_{\text{rnd}} < c_{\text{adm}}) dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} P(c_{\text{adm}} < T^{*}\mid\mathbf{X})P(c_{\text{adm}} < C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}) dF_{C_{\text{adm}}}(c_{\text{adm}})
                                            \end{aligned} \\
                                        &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}}} P(c_{\text{rnd}} < T^{*}\mid\mathbf{X})P(c_{\text{rnd}} < c_{\text{adm}}) dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} P(C_{\text{adm}} < T^{*}\mid\mathbf{X})P(C_{\text{adm}} < C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}) dF_{C_{\text{adm}}}(c_{\text{adm}})
                                            \end{aligned} \\
                                        &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) dF_{C_{\text{adm}}}(c_{\text{adm}})
                                            \end{aligned} \\
                                        &=: I_{\text{rnd}} + I_{\text{adm}}
                                                
\end{align*}
where $I_{\text{rnd}}$ describes the random censoring and $I_{\text{adm}}$ the administrative censoring.

Since $C_{\text{adm}} \in [\tau_{\text{adm}} - \tau_{\text{acc}}, \tau_{\text{adm}}]$ we have $P(c_{\text{rnd}} < C_{\text{adm}}) = S_{C_{\text{adm}}}(c_{\text{rnd}}) = 1$ for $c_{\text{rnd}} \in (0, \tau_{\text{adm}} - \tau_{\text{acc}})$. So we can rewrite $I_{\text{rnd}}$:
\begin{equation*}
    I_{\text{rnd}} = \int_{0}^{\tau_{\text{adm}} - \tau_{\text{acc}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) +
              \int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})
\end{equation*}

which gives us:
\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}} - \tau_{\text{acc}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) dF_{C_{\text{adm}}}(c_{\text{adm}})
                                            \end{aligned} \\
                                        &= \!\begin{aligned}[t]
                                                &\int_{0}^{\tau_{\text{adm}} - \tau_{\text{acc}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} + \\
                                                &\int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}})f_{C_{\text{adm}}}(c_{\text{adm}}) dc_{\text{adm}}
                                            \end{aligned} \\
                                        &= I_{\text{rnd,nadm}} + I_{\text{rnd,adm}} + I_{\text{adm}}
\end{align*}

The first integral $I_{\text{rnd,nadm}}$ describes the random censoring when no
administrative censoring can occur. The second integral $I_{\text{rnd,adm}}$ describes the random
censoring when also administrative censoring can occur and the third one $I_{\text{adm}}$ describes the administrative censoring.

Like in the case without accrual time, to get the probability of censoring in the simulated dataset, independent of
the covariates of a specific subject, we integrate $P(\delta=0\mid\mathbf{X},
\lambda_{C_{\text{rnd}}})$ over the covariate distribution $F_{X}$ and 
then the parameter $\lambda_{C_{\text{rnd}}}$ can be computed as a root finding
problem. 


### Edge cases

Lastly, we consider two edge cases, when $\tau_{\text{acc}} = 0$ and when $\tau_{\text{adm}} =
 \infty$. In the first case, the accrual time span is zero, i.e. all patients
 are accrued at or before the beginning of the study. So the general formula
 should collapse to the special formula we showed before.
If $\tau_{\text{acc}} = 0$ it follows that $C_{\text{adm}} \sim U([\tau_{\text{adm}}, \tau_{\text{adm}}])
= \delta_{\tau_{\text{adm}}}$, where $\delta_{\tau_{\text{adm}}}$ is the Dirac distribution
at $\tau_{\text{adm}}$. It follows:

\begin{align*}
    I_{\text{adm}} &= \int_{\tau_{\text{adm}} - \tau_{\text{acc}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) dF_{C_{\text{adm}}}(c_{\text{adm}}) \\
    &= \int_{\tau_{\text{adm}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) d\delta_{\tau_{\text{adm}}}(c_{\text{adm}}) \\
    &= \int_{\{\tau_{\text{adm}}\}} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) d\delta_{\tau_{\text{adm}}}(c_{\text{adm}}) \\
    &= S_{T^{*}\mid\mathbf{X}}(\tau_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(\tau_{\text{adm}})
\end{align*}

For $I_{\text{rnd,adm}}$ we have:
\begin{align*}
    I_{\text{rnd,adm}} = \int_{\tau_{\text{adm}}}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c)S_{C_{\text{adm}}}(c) dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c) = \int_{\{\tau_{\text{adm}}\}} S_{T^{*}\mid\mathbf{X}}(c)S_{C_{\text{adm}}}(c) dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c) = 0
\end{align*}
because $F_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}$ is a continuous distribution function (hence, a continuous measure), so the integral over one
point is zero.

So when $\tau_{\text{acc}} = 0$ we get: 
\begin{equation*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) = \int_{0}^{\tau_{\text{adm}}} S_{T^{*}\mid\mathbf{X}}(c)f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c)dc + S_{T^{*}\mid\mathbf{X}}(\tau_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(\tau_{\text{adm}})
\end{equation*}
which is the special case we saw in the [Censoring without accrual](#sec:no_accrual).

In the second case, the administrative censoring time is infinite, i.e.
patients are observed until the event or random censoring occurs. This results
in three improper integrals in the general formula. We can solve these by
considering the limit of these integrals as $\tau_{\text{adm}} \to \infty$. First we
take a look at $I_{\text{rnd}}$.
If $\tau_{\text{adm}} \to \infty$, then $\tau_{\text{adm}}-\tau_{\text{acc}} \to \infty$. So we have:

\begin{align*}
    I_{\text{rnd,adm}} &= \int_{\infty}^{\infty} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) \\
    &= \lim_{a \to \infty} \int_{a}^{a} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})dF_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}}) \\
    &= \lim_{a \to \infty} 0 = 0.
\end{align*}

Next, we consider $I_{\text{adm}}$. If $\tau_{\text{adm}} \to \infty$, then
$C_{\text{adm}} \sim U([\infty, \infty]) = \delta_{\infty} = \lim_{a \to \infty}
\delta_{a}$. It follows:

\begin{align*}
    I_{\text{adm}} &= \int_{\infty}^{\infty} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) dF_{C_{\text{adm}}}(c_{\text{adm}}) \\
    &= \lim_{a \to \infty} \int_{a}^{a} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}}) d\delta_{a}(c_{\text{adm}}) \\
    &= \lim_{a \to \infty} S_{T^{*}\mid\mathbf{X}}(a)S_{    C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(a) = 0.
\end{align*}

Note that $\lim_{a \to \infty} S_{X}(a) = 0$ for every continuous survival
function $S_{X}$.

So, when $\tau_{\text{adm}} \to \infty$ the general formula collapses to:
\begin{equation*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) = \int_{0}^{\infty} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}}
\end{equation*}

We see when $\tau_{\text{adm}} \to \infty$, the accrual time does not affect
the calculation of the censoring parameter. In other words, accrual time becomes irrelevant because each
patient is observed until the event or random censoring.

## Basic Usage

```{r, basic_estimator}
library(CensRCalc)
library(progressr) # For progress bar during root finding
library(logger) # For logging messages

tau_acc <- 1
tau_adm <- 3


event_surv <- function(t, x) 1 - punif(t, min = 0, max = x)
cens_surv <- function(t, lambda_c) 1 - punif(t, min = 0, max = lambda_c)
cens_dens <- function(t, lambda_c) dunif(t, min = 0, max = lambda_c)
cov_dens <- function(x) dunif(x, min = 1, max = 2)
cov_bounds <- list(x = c(1, 2))
bounds <- c(0.1, 10)

log_threshold(DEBUG, namespace = "CensRCalc") # Set logger to DEBUG level
estimator <- estimate_cens_prop(event_surv, cov_dens, cov_bounds,
  time_accrual = tau_acc,
  time_admin_cens = tau_adm,
  cens_surv = cens_surv,
  cens_dens = cens_dens,
  target_bounds = bounds
)
log_threshold(INFO, namespace = "CensRCalc") # Reset logger to INFO level

lambda_c_to_plot <- seq(bounds[1], bounds[2], length.out = 100)
cens_prop <- sapply(lambda_c_to_plot, estimator)
plot(lambda_c_to_plot, cens_prop,
  type = "l", xlab = expression(lambda[C]),
  ylab = "Censoring Proportion"
)
```

```{r, find_param}
p_c <- 0.3 # Target censoring proportion

with_progress({
  result <- find_cens_param(p_c, event_surv, cov_dens, cov_bounds,
    time_accrual = tau_acc,
    time_admin_cens = tau_adm,
    cens_surv = cens_surv,
    cens_dens = cens_dens,
    target_bounds = bounds
  )
  print(result)
})

lambda_c <- result$parameter
```

```{r, simulate_data}
set.seed(123)

n <- 10000
x <- runif(n, min = 1, max = 2) # Covariate
t_true <- runif(n, min = 0, max = x) # Event times uniform (0, X)
c_rnd <- runif(n, min = 0, max = lambda_c) # Random censoring times uniform
r <- runif(n, min = 0, max = tau_acc) # Recruitment times uniform (0, tau_acc)
c_adm <- tau_adm - r # Administrative follow-up after recruitment
c <- pmin(c_rnd, c_adm)
t_obs <- pmin(t_true, c)
delta <- as.integer(t_true <= c)

mean(delta == 0) # Empirical censoring proportion
```


## Analytical Background of the Example

For this, we use the simplest distributions for the event and censoring times, 
i.e. uniform distributions:
\begin{align*}
    T^{*}\mid\mathbf{X} &\sim \text{U}(0, \lambda_{T^{*}}) \\
    C_{\text{rnd}} &\sim \text{U}(0, \lambda_{C_{\text{rnd}}})
\end{align*}
where $\lambda_{T^{*}} = \beta_0 + \beta_1X$. Let $\beta_0 = 0$, $\beta_1 = 1$
and $X \sim \text{U}(1,2)$.

Let $\tau_{\text{acc}} = 1$, i.e. patients are accrued within a one year time span.
The special case is not shown because there is nothing to be aware of when
setting $\tau_{\text{acc}} = 0$. Furthermore, $\tau_{\text{adm}} = 3$, i.e. patients are 
observed for three years. That means $C_{\text{adm}} \sim U([2, 3])$.
The target value of the expected censoring proportion is $p_{C} = 0.3$.
The goal is to calculate the parameter $\lambda_{C_{\text{rnd}}}$ of the random
censoring distribution so that the expected censoring proportion in a
simulated dataset is $p_{C}$.

For calculating the censoring probability for a specific subject, we need the
survival functions of the event time, random censoring time and the
administrative censoring time as well as the density functions of the random
and administrative censoring time. The survival functions are given by:

\begin{align*}
    S_{T^{*}\mid\mathbf{X}}(t) &= 1 - F_{T^{*}}(t) = 1 - \frac{t-0}{X-0} = \frac{X-t}{X} \\
    S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(t) &= 1 - F_{C_{\text{rnd}}}(t) = 1 - \frac{t-0}{\lambda_{C_{rnd}}-0} = \frac{\lambda_{C_{rnd}}-t}{\lambda_{C_{rnd}}} \\
    S_{C_{\text{adm}}}(t) &= 1 - F_{C_{\text{adm}}}(t) = 1 - \frac{t - (\tau_{\text{adm}} - \tau_{\text{acc}})}{\tau_{\text{adm}} - (\tau_{\text{adm}} - \tau_{\text{acc}})} = 1 - \frac{t - 2}{1} = 3 - t
\end{align*}

The density functions are given by:
\begin{align*}
    f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(t) &= \frac{1}{\lambda_{C_{rnd}}-0} = \frac{1}{\lambda_{C_{rnd}}} \\
    f_{C_{\text{adm}}}(t) &= \frac{1}{\tau_{\text{adm}} - (\tau_{\text{adm}} - \tau_{\text{acc}})} = 1
\end{align*}

The probability is then calculated as follows:
\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= \!\begin{aligned}[t]
                                                &\int_{0}^{2} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} + \\
                                                &\int_{2}^{3} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})S_{C_{\text{adm}}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} + \\
                                                &\int_{2}^{3} S_{T^{*}\mid\mathbf{X}}(c_{\text{adm}})S_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{adm}})f_{C_{\text{adm}}}(c_{\text{adm}}) dc_{\text{adm}}
                                            \end{aligned}
\end{align*}

Since the support of $T^*$ is $[0,X]$ with $X \in [1,2]$, the survival function $S_{T^{*}\mid\mathbf{X}}(t) = 0$ for all $t \geq 2$. 
Thus, the second and third integrals, which integrate over $[2,3]$, are zero. The first integral has to be split up into two integrals:

\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= \int_{0}^{2} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} \\
    &= \int_{0}^{X} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} + \int_{X}^{2} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} \\
\end{align*}

Using the same reason as above, the second integral over $[X,2]$, is also zero. Thus, we have:

\begin{align*}
    P(\delta=0\mid\mathbf{X}, \lambda_{C_{\text{rnd}}}) &= \int_{0}^{X} S_{T^{*}\mid\mathbf{X}}(c_{\text{rnd}})f_{C_{\text{rnd}}\mid\lambda_{C_{\text{rnd}}}}(c_{\text{rnd}})dc_{\text{rnd}} \\
    &= \int_{0}^{X} \frac{X - c_{\text{rnd}}}{X} \cdot \frac{1}{\lambda_{C_{rnd}}} dc_{\text{rnd}} \\
    &= \frac{1}{\lambda_{C_{rnd}} X} \int_{0}^{X} (X - c_{\text{rnd}}) dc_{\text{rnd}} \\
    &= \frac{1}{\lambda_{C_{rnd}} X} \left[ X c_{\text{rnd}} - \frac{c_{\text{rnd}}^2}{2} \right]_{0}^{X} \\
    &= \frac{1}{\lambda_{C_{rnd}} X} \left( X^2 - \frac{X^2}{2} \right) \\
    &= \frac{X}{2 \lambda_{C_{rnd}}}
\end{align*}

Next we calculate the expected censoring proportion in the simulated 
dataset, independent of the covariates of a specific subject:

\begin{align*}
    P(\delta=0\mid\lambda_{C}) &= E_X\left[P(\delta=0\mid\mathbf{X}, \lambda_{C})\right] \\
    &= \int_{D_{\mathbf{X}}} P(\delta=0\mid\mathbf{X}=x, \lambda_{C_{rnd}})f_{X}(x)dx \\
    &= \int_{1}^{2} \frac{X}{2 \lambda_{C_{rnd}}}f_{X}(x)dx \\
    &= \int_{1}^{2} \frac{X}{2 \lambda_{C_{rnd}}} \cdot 1 dx \\
    &= \frac{1}{2 \lambda_{C_{rnd}}} \int_{1}^{2}X dx \\
    &= \frac{1}{2 \lambda_{C_{rnd}}} \left[\frac{X^2}{2} \right]_{1}^{2} \\
    &= \frac{1}{4 \lambda_{C_{rnd}}} \left(4 - 1\right) \\
    &= \frac{3}{4 \lambda_{C_{rnd}}}
\end{align*}

In the last step we can calculate the parameter $\lambda_{C_{rnd}}$:
\begin{align*}
    &P(\delta=0\mid\lambda_{C_{\text{rnd}}}) \overset{!}{=} 0.3 \\
    &\leftrightarrow \frac{3}{4 \lambda_{C_{rnd}}} = 0.3 \\
    &\leftrightarrow \frac{3}{4 \cdot 0.3} = \lambda_{C_{rnd}} \\
    &\leftrightarrow \lambda_{C_{rnd}} = \frac{30}{12} = 2.5
\end{align*}

## Session Info

```{r session_info}
sessionInfo()
```

## References
